Towards a free/open-source universal-dependency treebank for Kazakh

### 2015-06-21 2015-turklang-kazdep/paper.tex
kazakh section done?
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60940 72bbbca6-d526-0410-a7d9-f06f51895060

Kazakh, like most Turkic languages, makes frequent use of non-finite verb forms by deriving verbal adjectives, verbal nouns, and verbal adverbs from verbs.  Verbal adjective phrases modify nouns, as in \kazakh{мені \textbf{көрген} дәрігер} \gloss{the doctor that saw me}.  Verbal nouns can function as subjects or complements of verbs or copulas, as in \kazakh{Дәрігер сені \textbf{көргенін} білген жоқпын.} \gloss{I didn't know that the doctor had seen you.} and \kazakh{Дәрігер сені \textbf{көргені} жақсы болыпты.} \gloss{It's good that doctor saw you.}

Another category that may also be non-overt is the copula.  The primary strategy for copula constructions in Kazakh is the use of a defective verb \kazakh{е-}.  In the present tense, the verb itself does not surface, but agreement morphology surfaces (cliticised to the previous word) in all but the third person forms (e.g., \kazakh{Мен үйде\textbf{мін}.} \gloss{I'm at home.} versus \kazakh{Ол үйде.} \gloss{S/he's at home.}).  The defective copula verb also surfaces in the recent past tense (e.g., \kazakh{Мен үйде еді\textbf{м}.} \gloss{I was at home.}, \kazakh{Інім үйде еді.} \gloss{My younger brother was at home.}).  Of particular interest are non-finite forms of the copula.  Copula clauses can be attributive, as in \kazakh{үйі жақсы дәрігер} \gloss{the doctor who has a nice house}, or literally \gloss{[his/her house is nice] the doctor}.  While this construction never has overt copula marking, copula clauses which are complements of verbs always have an overt copula form, e.g. \kazakh{Дәрігердің үйі жақсы \textbf{екенін} білген жоқпын.} \gloss{I didn't know that the doctor's house was nice.}  Here, \kazakh{екен} is a suppletive gerundive form of the copula and is marked as accusative.

>>>>

Kazakh, like most Turkic languages, makes frequent use of non-finite verb forms by deriving verbal adjectives, verbal nouns, and verbal adverbs from verbs.  Verbal adjective phrases modify nouns, as in \kazakh{\textbf{мені көрген} дәрігер} \gloss{the doctor \textbf{that saw me}}.  Verbal nouns can function as subjects or complements of verbs or copulas, as in \kazakh{Дәрігер сені \textbf{көргенін} білген жоқпын.} \gloss{I didn't know that the doctor had seen you.} and \kazakh{Дәрігер сені \textbf{көргені} жақсы болыпты.} \gloss{It's good that doctor saw you.}  Verbal adverbs allow a verb phrase to function as a clausal verbal adjunct, as in \kazakh{Мен \textbf{дәрігерді көріп} қуанып кеттім.} \gloss{\textbf{Seeing the doctor}, I got happy.}  Verbal nouns with certain case morphology and/or occurring with certain postpositions can also function as a verbal adjunct much in the same way that verbal adverb clauses do, as in \kazakh{Мен \textbf{дәрігерді көргенде} қуанып кеттім.} \gloss{I got happy \textbf{when I saw the doctor}.}, here with a verbal noun in the locative case.

Another category that may also be non-overt is the copula.  The primary strategy for copula constructions in Kazakh is the use of a defective verb \kazakh{е-}.  In the present tense, the verb itself does not surface, but agreement morphology surfaces (cliticised to the previous word) in all but the third person forms (e.g., \kazakh{Мен үйде\textbf{мін}.} \gloss{I'm at home.} versus \kazakh{Ол үйде.} \gloss{S/he's at home.}).  The defective copula verb also surfaces in the recent past tense (e.g., \kazakh{Мен үйде еді\textbf{м}.} \gloss{I was at home.}, \kazakh{Інім үйде еді.} \gloss{My younger brother was at home.}).  Of particular interest are non-finite forms of the copula.  Copula clauses can be attributive, as in \kazakh{\textbf{үйі жақсы} дәрігер} \gloss{the doctor \textbf{who has a nice house}}, or literally \gloss{\textbf{his/her house is nice} the doctor}.  While this construction never has overt copula marking, copula clauses which are complements of verbs always have an overt copula form, e.g. \kazakh{Дәрігердің үйі жақсы \textbf{екенін} білген жоқпын.} \gloss{I didn't know \textbf{that} the doctor's house \textbf{was} nice.}  Here, \kazakh{екен} is a suppletive gerundive form of the copula and is marked as accusative.
